Cartogram Mapping and its Application to Cancer Data Visualisation

### 2019-09-03 review_article.Rmd
Figure captions

---
title: "Cartogram Mapping and its Application to Cancer Data Visualisation"
author: "Stephanie Kobakian, Jessie Roberts and Dianne Cook"
bibliography: reviewbib.bib
output:
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
fontsize: 11pt
link-citations: yes
fig_caption: yes
biblio-style: vancouver
yolo: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
```
<!--

@@

```{r choro, fig.cap = "A choropleth map of the United States of America. Each state has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. The state of Utah has an unusually low rate of lung and bronchus incidence. The state of Kentucky is high, the surrounding states also have high rates. There is a pattern of average higher rates toward the East Coast and lower rates toward the West Coast.", message=FALSE, warning=FALSE, fig.align='center', out.width='80%'}
knitr::include_graphics("figures/ggchoro.png")
```

@@

library(grid)
library(png)
library(gridExtra)

plot1 <- png::readPNG('cancer_map_images/choropleths/3.choropleth.png')
plot2 <- png::readPNG('cancer_map_images/choropleths/15.choropleth.png')
plot3 <- png::readPNG('cancer_map_images/choropleths/25.choropleth.png')
plot4 <- png::readPNG('cancer_map_images/choropleths/20.choropleth1.png')
plot5 <- png::readPNG('cancer_map_images/choropleths/26.choropleth.png')
plot6 <- png::readPNG('cancer_map_images/choropleths/14.choropleth2.png')
plot7 <- png::readPNG('cancer_map_images/choropleths/21.choropleth.png')
plot8 <- png::readPNG('cancer_map_images/choropleths/23.choropleth.png')
plot9 <- png::readPNG('cancer_map_images/choropleths/7.choropleth.png')
plot10 <- png::readPNG('cancer_map_images/choropleths/30.choropleth.png')
plot11 <- png::readPNG('cancer_map_images/choropleths/29.choropleth.png')
plot12 <- png::readPNG('cancer_map_images/choropleths/1.png')
plot13 <- png::readPNG('cancer_map_images/choropleths/24.choropleth.png')
plot14 <- png::readPNG('cancer_map_images/choropleths/6.choropleth.png')
plot15 <- png::readPNG('cancer_map_images/choropleths/27.choropleth.png')
plot16 <- png::readPNG('cancer_map_images/choropleths/8.choropleth.png')
plot17 <- png::readPNG('cancer_map_images/choropleths/5.choropleth.png')

@@

```{r choropleth-grid-print, echo = FALSE, results = "asis"}

#knitr::include_graphics("figures/choropleth_grid.png")
```

@@

Cancer maps are effective tools for communicating to wide range of audiences, including the general public and others not trained in statistical analyses. These visualisation enable these non-expert audiences to access the outputs of sophisticated statistical analyses. Cruickshank's (1947) as cited by @DMAHP, discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly cancer maps utilised to communicate to the public and other non-expert audiences are choropleths. A grey literature review of cancer atlases available in the public domain [@roberts2019communication] identified 33 cancer atlases published on the internet between January 2010 and November 2015, all of which were choropleths. These maps were published by non-commercial organisations, including not-for-profits, government, research organisations, advocacy groups or government funded partnerships.Only one map was published by a commercial entity.  The use of choropleths within the public domain mirrors the heavy use of choropleth maps within the research literature, discussed above.

<!-- geographic coverage-->
<!-- is this that interesting to the reader-->
The cancer atlases covered geographies from all around the world, but only four were global atlases. Most focused on single nations, the United States was considered by eleven atlases, the United Kingdom by seven, followed by three of Australia, two of Canada, and one of each from Switzerland, Germany, Norway. One atlas covered the European Union.
Not all maps had a national focus and ten covered a region or state rather than an entire nation. The states or counties/regions covered were South Australia (AUS), Queensland (AUS), Ontario (CAN), Valencia (Spain), Pennsylvania county Massachusetts (US), New Hampshire (US), Cape Cod (US), Missouri (US), Florida (US), New York State (US) and Arizona (US). Figure 2.a.b and c show a global, national and state choropleth respectively.

@@

<!-- insert figure with 3 examples of supplementary graphs-->

@@

```{r choroCRS, fig.cap = "Four choropleth maps of the United States of America using various coordinate reference systems. Each state has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. The map projections alter the shapes and angles of the boundaries of each state. Maps a and b are similar in their straight edges, unlike maps c and d which curve on the northern United States border. ", message=FALSE, warning=FALSE, fig.align='center', out.width='80%'}
knitr::include_graphics("figures/ggchoroCRS.png")
```

@@

```{r ggcont, fig.cap = "A contiguous cartogram map of the United States of America. Each state has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. Each stated shape has been distorted according to the population of the state in 2015. The state of California has become much larger due to it's large population density. This draws attention to the densely populated North East region, and detracts from the Mid West.", message=FALSE, warning=FALSE, fig.align='center', out.width='80%'}
knitr::include_graphics("figures/ggcont.png")
```

@@

```{r ggncont, fig.cap = "A Non - contiguous cartogram map of the United States of America. Each state has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. Each state shape has been maintained, but the size has altered according to the population of the state in 2015. The state of California has remained closer to it's original size than it's surrounding states. The North East states have remained closer to their geographical size, in the case of Massachusetts and Connecticut. This draws attention to the densely populated North East region, and the sparse Mid West.", message=FALSE, warning=FALSE, fig.align='center', out.width='80%'}
knitr::include_graphics("figures/ggncont.png")
```

@@

```{r ggdorl, fig.cap = "A dorling cartogram of the Unites States of America. Each state has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. Each state has been represented by a circle, but the size was determined by the population of the state in 2015. The North East states remain closer to their neighbours, and may be displaced from their geographic location. The sparsity of the population in the Mid West is highlight by the distance between the circles, located at the geographic centroids.", message=FALSE, warning=FALSE, fig.align='center', out.width='80%'}
knitr::include_graphics("figures/ggdorl.png")
```

@@

Square grids have been used by FiveThirtyEight, Bloomberg Business, The Guardian, The Washington Post, The New York Times and NPR. Where each area is represented by a square of the same dimensions.
An example is provided in \ref{fig:ggtilegram}

```{r ggtilegram, fig.cap = "A tilegram of the Unites States of America. Each state has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. Each state has been represented by a square, each square is the same size. The neighbouring states are easily contrasted, however the north east regions has been displaced from their geographic location. The sparsity of the population in the Mid West is highlight by the light yellow colour, the age adjusted rate in Kentucky is the darkest and its neighbours are similar..", message=FALSE, warning=FALSE, fig.align='center', out.width='80%'}
knitr::include_graphics("figures/ggtilegram.png")
```

@@

|  | Choropleth | Contiguous Cartogram | Non-contiguous Cartogram | Dorling | Tilegrams |
|:---------------------------------------------------:|------------|----------------------|--------------------------|---------|-----------|
| Feature preservation: |  |  |  |  |  |
| Preserves shapes | Y | N | Y | N | N |
| Preserves neighbours | Y | Y | Y | S | S |
| Uniform use of shape | N | N | N | Y | Y |
| Statistical transformation of the base map: |  |  |  |  |  |
| Used for bi-variate maps eg. population & statistic | N | Y | Y | Y | Y |

@@

It was in the development of this online cancer atlas that methods for disease map displays, and visual communication strategies were explored. We are thankful for the opportunity to write about these visualisation displays and the situations in which they are appropriate.

>>>>

---
title: "Cartogram Mapping and its Application to Cancer Data Visualisation"
author: "Stephanie Kobakian, Jessie Roberts and Dianne Cook"
bibliography: reviewbib.bib
output:
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
    fig_caption: true
fontsize: 11pt
link-citations: yes
biblio-style: vancouver
yolo: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
library(ggplot2)
library(grid)
library(png)
library(gridExtra)
```
<!--

@@

```{r choro, fig.cap = "A choropleth map of the United States of America. Each state has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. The state of Utah has an unusually low rate of lung and bronchus incidence. The state of Kentucky is high, the surrounding states also have high rates. There is a pattern of average higher rates toward the East Coast and lower rates toward the West Coast.", message=FALSE, warning=FALSE, out.width='80%'}
knitr::include_graphics("figures/ggchoro.png")
```

@@

plot1 <- png::readPNG('cancer_map_images/choropleths/3.choropleth.png')
plot2 <- png::readPNG('cancer_map_images/choropleths/15.choropleth.png')
plot3 <- png::readPNG('cancer_map_images/choropleths/25.choropleth.png')
plot4 <- png::readPNG('cancer_map_images/choropleths/20.choropleth1.png')
plot5 <- png::readPNG('cancer_map_images/choropleths/26.choropleth.png')
plot6 <- png::readPNG('cancer_map_images/choropleths/14.choropleth2.png')
plot7 <- png::readPNG('cancer_map_images/choropleths/21.choropleth.png')
plot8 <- png::readPNG('cancer_map_images/choropleths/23.choropleth.png')
plot9 <- png::readPNG('cancer_map_images/choropleths/7.choropleth.png')

@@

@@

Cancer maps are effective tools for communicating to wide range of audiences, including the general public and others not trained in statistical analyses. These visualisation enable thse non-expert audiences to access the outputs of sophisticated statistical analyses. Cruickshank's (1947) as cited by @DMAHP, discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly cancer maps utilised to communicate to the public and other non-expert audiences are choropleths. A grey literature review of cancer atlases available in the public domain [@roberts2019communication] identified 33 cancer atlases published on the internet between January 2010 and November 2015, all of which were choropleths. These maps were published by non-commercial organisations, including not-for-profits, government, research organisations, advocacy groups or government funded partnerships.Only one map was published by a commercial entity.  The use of choropleths within the public domain mirrors the heavy use of choropleth maps within the research literature, discussed above. 
The cancer atlases covered geographies from all around the world. One atlas covered the European Union. Most focused on single nations, the United States was considered by eleven atlases, the United Kingdom by seven, followed by three of Australia, two of Canada, and one of each from Switzerland, Germany, Norway. 
Figure 2.a.b and c show a global, national and state choropleth respectively. 
Not all maps had a national focus and ten covered a region or state rather than an entire nation. The states or counties/regions covered were South Australia (AUS), Queensland (AUS), Ontario (CAN), Valencia (Spain), Pennsylvania county Massachusetts (US), New Hampshire (US), Cape Cod (US), Missouri (US), Florida (US), New York State (US) and Arizona (US).

@@

```{r supp-grid-create, fig.cap = "Interactive controls of displays in public facing choropleth cancer maps. From left to right and top to bottom: [a. Globocan 2018: Cancer Today](http://globocan.iarc.fr/Pages/Map.aspx)(Controls for indicator, sex, age groups, continents, and cancer types) , [b. Bowel Cancer Australia Atlas](http://www.bowelcanceratlas.org/) (Controls for indicator, age, year, and geographical areas), [c. Global Cancer Map](http://globalcancermap.com/) (Controls for indicator, and cancer type), [d. United States Cancer Statistics: Data Visualizations](https://gis.cdc.gov/Cancer/USCS/DataViz.html) (Controls for demographics, trends, geography, indicators, risk factors, sex, cancer type, and year)."}
i3 <- png::readPNG('cancer_map_images/interactivity/3.interactivity.png')
i15 <- png::readPNG('cancer_map_images/interactivity/15.interactivity.png')
i4 <- png::readPNG('cancer_map_images/interactivity/4.interactivity.png')
i5 <- png::readPNG('cancer_map_images/interactivity/5.interactivity.png')
i7 <- png::readPNG('cancer_map_images/interactivity/7.interactivity.png')

lay <- rbind(c(1,1,2,2,2,2),
             c(1,1,2,2,2,2),
             c(1,1,2,2,2,2),
             c(1,1,3,3,3,3),
             c(1,1,4,4,4,4),
             c(1,1,5,5,5,5))

supp_grid <- gridExtra::grid.arrange(
  grid::rasterGrob(i3), grid::rasterGrob(i15),
  grid::rasterGrob(i5), grid::rasterGrob(i7), grid::rasterGrob(i4), layout_matrix = lay)

ggsave(filename = "figures/supp_grid.png", plot = supp_grid, device = "png", width = 6, height = 5, dpi = 300)
```

@@

```{r choroCRS, fig.cap = "Four choropleth maps of the United States of America using various coordinate reference systems. Each state has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. The map projections alter the shapes and angles of the boundaries of each state. Maps a and b are similar in their straight edges, unlike maps c and d which curve on the northern United States border.", message=FALSE, warning=FALSE, out.width='80%'}
knitr::include_graphics("figures/ggchoroCRS.png")
```

@@

```{r ggcont, fig.cap = "A contiguous cartogram map of the United States of America. Each state has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. Each stated shape has been distorted according to the population of the state in 2015. The state of California has become much larger due to it's large population density. This draws attention to the densely populated North East region, and detracts from the Mid West.", message=FALSE, warning=FALSE, out.width='80%'}
knitr::include_graphics("figures/ggcont.png")
```

@@

```{r ggncont, fig.cap = "A Non - contiguous cartogram map of the United States of America. Each state has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. Each state shape has been maintained, but the size has altered according to the population of the state in 2015. The state of California has remained closer to it's original size than it's surrounding states. The North East states have remained closer to their geographical size, in the case of Massachusetts and Connecticut. This draws attention to the densely populated North East region, and the sparse Mid West.", message=FALSE, warning=FALSE, out.width='80%'}
knitr::include_graphics("figures/ggncont.png")
```

@@

```{r ggdorl, fig.cap = "A dorling cartogram of the Unites States of America. Each state has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. Each state has been represented by a circle, but the size was determined by the population of the state in 2015. The North East states remain closer to their neighbours, and may be displaced from their geographic location. The sparsity of the population in the Mid West is highlight by the distance between the circles, located at the geographic centroids.", message=FALSE, warning=FALSE, out.width='80%'}
knitr::include_graphics("figures/ggdorl.png")
```

@@

Square grids have been used by FiveThirtyEight, Bloomberg Business, The Guardian, The Washington Post, The New York Times and NPR. Where each area is represented by a square of the same dimensions.
Figure \ref{fig:ggtilegram} shows a square tilegram representing each state using a single square.

```{r ggtilegram, fig.cap = "A tilegram of the Unites States of America. Each state has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. Each state has been represented by a square, each square is the same size. The neighbouring states are easily contrasted, however the north east regions has been displaced from their geographic location. The sparsity of the population in the Mid West is highlight by the light yellow colour, the age adjusted rate in Kentucky is the darkest and its neighbours are similar.", message=FALSE, warning=FALSE, out.width='80%'}
knitr::include_graphics("figures/ggtilegram.png")
```

@@

|  | Choropleth | Contiguous | Non-contiguous | Dorling | Tilegrams |
|:-----------:|------------|-----------|---------------|---------|-----------|
| Feature preservation: |  |  |  |  |  |
| Preserves shapes | Y | N | Y | N | N |
| Preserves neighbours | Y | Y | Y | S | S |
| Uniform use of shape | N | N | N | Y | Y |
| Bivariate display* | N | Y | Y | Y | Y |
* A bivariate display refers to a map that allows two variables to be plotted. One variable via the size of the area, and another by filling the area with a colour.

@@

It was in the development of this online cancer atlas that methods for disease map displays, and visual communication strategies were explored. We are thankful for the opportunity to write about these visualisations and the situations in which they are appropriate.
